---
title: "TimeSeries_KevinUpdates"
author: "Zoe Zilz"
date: "5/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(tidyverse)
```
## Updates From Meeting With Kevin  
### Include:  
- Consider subsuming algae as equilibrium value A* into grazer equation  
- **Grazers (snails) need to be limited by density dependent mortality ** 
  -- does this add on top of normal mortality, or replace it?  
- **Fix grazer equation so they become saturated by algae at a rate A/h+A, with h being handling time  **
  -- same for both infected and uninfected  
- Change cercaria infection parameter to either be it's own dynamic equation or use separation of time scales to subsume
  -- make sure cercariae are produced as a function of algae consumed by grazers
- add a non-dynamic alternative prey for definitive hosts

```{r set parameters, include = FALSE}

source("kevinmods_paramvals.R") # continually edited as chunks of the model are added

# should I also make a file for initial state variable values?
## yes

source("build_statevars0.R")

tset <- seq(from=0, to=1000, length.out = 5000) # definitely going to have to tweak time scale later

# color for algae
Acol <- "green"
# color for 1IH snails
Fcol <- "orange"
# color for 2IH fish
Scol <- "purple"
# color for DH birbs
Dcol <- "blue"
# color for parasite eggs
Pcol <- "red"


```


## One For-Loop at a time
### Let's Start with algae
Kevin says algae carrying capacity is i/d_A (duh)
He also says we could subsume A* into the rest of the equations, but I don't want to do that yet

$$
\begin{align}
\newline
\frac{d A}{dt}  &= i - A(d_A + \gamma_I F_I  + \gamma_U F_U) \\ 
\end{align}
$$

```{r algae_only}
# initial values for FI and FU are zero
F_I <- 0
F_U <- 0

# set up empty vectors for algae A
A.simu <- NaN*tset; A.simu[1] <- A0

# for loop
for(i in 2:length(tset)){
	dt <- tset[i]-tset[i-1]
	A <- A.simu[i-1]
	dA <- (inp - A*(d_A + gamma_I*F_I + gamma_U*F_U))*dt
	A.simu[i] <- A + dA
}

# plot

plot(tset, A.simu,type='l',las=1,lwd=2,ylab='Population Size',xlab='Time', col=Acol,ylim=c(0,100))
legend(x=70,y=90,legend=c('Algae'),lwd=2,col=c(Acol))
```

ok, that worked, algae is just growing logistically due to death?

### Now lets add in first intermediate hosts (snails) but without parasites in the system
We should see consumption of algae and growth of snails (I didnt set this up to have a carrying capacity or predator but I probably should...)
Kevin thinks adding in density dependence for snails (cannibalism and shelter limitation) by incorporating -mF^2 (m = density dependent mortality)
also adding handling time for snails eating algae (A/(h+A))
turn castration on by setting eF_I = 0
$$
\begin{align}
\newline
\frac{d F_U}{dt} &= F_U (\gamma_U e_{FU}\frac{A}{h_F+A} - d_{FU}) -P\beta_I F_U + F_I(\gamma_I e_{FI}\frac{A}{h_F+A}) - m{F_U}^2\\
\newline
\frac{d F_I}{dt} &= P\beta_I F_U - d_{FI}F_I - m{F_I}^2\\
\end{align}
$$
```{r algae_and_1IH}
# we're not adding in parasites yet so
P <- 0

# set up empty vectors and initial values
A.simu <- NaN*tset; A.simu[1] <- A0
FI.simu <- NaN*tset; FI.simu[1] <- 0 # no parasites yet so this is 0
FU.simu <- NaN*tset; FU.simu[1] <- F_U0

# for loop
for(i in 2:length(tset)){
	dt <- tset[i]-tset[i-1]
	A <- A.simu[i-1]
	F_U <- FU.simu[i-1]
	F_I <- FI.simu[i-1]
	# algae
	dA <- (inp - A*(d_A + gamma_I*F_I + gamma_U*F_U))*dt
	# uninfected
	dF_U <- (F_U*(gamma_U*e_FU*(A/(h_F +A)) - d_FU)  # grazing * conversion efficiency * algae pop/handling time - death
	         - (P*beta_I*F_U)                   # parasites (0) * infection * uninfected pop, should = 0
	         + F_I*(gamma_I*e_FI*(A/(h_F +A)))  # births from infecteds, should = 0
	         - (m_FU * F_U)^2)*dt               # density dependent mortality
	# infected
	dF_I <- ((P*beta_I*F_U) 
	         - (d_FI*F_I)                       # deaths
	         - (m_FI * F_I)^2)*dt               # density dependent mortality
	A.simu[i] <- A + dA
	FU.simu[i] <- F_U + dF_U
	FI.simu[i] <- F_I + dF_I
}

# plot

plot(tset, A.simu,type='l',las=1,lwd=2,ylab='Population Size',xlab='Time', col=Acol,ylim=c(0,20))
lines(tset, FU.simu, type = 'l', las = 1, lwd = 2, col = Fcol)
lines(tset, FI.simu, type = 'l', las = 1, lty = 3, col = Fcol)
legend(x=70,y=90,legend=c('Algae', '1st Intermediate Host'),lwd=2,col=c(Acol, Fcol))
```

Looks good so far, reaching an equilibrium. Here we have both natural mortality and density dependent mortality, not sure how appropriate that is, **but check with group.**

### Let's add second intermediate hosts
We should see fish coming out of nowhere (might need to fix this) and not being eaten yet so they're gonna explode, snails are still not being eaten or infected so they will stay same, and algae should stay the same. not much should change
$$
\begin{align}
\frac{d S_U}{dt} &= r_{SU}S_U - (D_U e_{DU} S_U \alpha_{DUSU} + D_Ie_{DI}S_U \alpha_{DISU}) + r_{SI}S_I - F_I S_U c\\
\newline
\frac{d S_I}{dt} &= F_I S_U c - (D_U e_{DU}\alpha_{DUSI} + D_Ie_{DI}\alpha_{DISI})\\
\newline
\end{align}
$$
```{r algae_1IH_2IH}
# we're not adding in parasites yet so
P <- 0
c <- 0

# no predators yet either so
D_U <- 0
D_I <- 0


# set up empty vectors for algae A
A.simu <- NaN*tset; A.simu[1] <- A0
FI.simu <- NaN*tset; FI.simu[1] <- 0 # no parasites yet so this is 0
FU.simu <- NaN*tset; FU.simu[1] <- F_U0
SI.simu <- NaN*tset; SI.simu[1] <- 0 # no parasites yet so this is 0
SU.simu <- NaN*tset; SU.simu[1] <- S_U0


# for loop
for(i in 2:length(tset)){
	dt <- tset[i]-tset[i-1]
	A <- A.simu[i-1]
	F_U <- FU.simu[i-1]
	F_I <- FI.simu[i-1]
	S_U <- SU.simu[i-1]
	S_I <- SI.simu[i-1]
	# algae
	dA <- (inp - A*(d_A + gamma_I*F_I + gamma_U*F_U))*dt
	# uninfected 1IH
	dF_U <- (F_U*(gamma_U*e_FU*(A/(h_F +A)) - d_FU)  # grazing * conversion efficiency * algae pop/handling time - death
	         - (P*beta_I*F_U)                   # parasites (0) * infection * uninfected pop, should = 0
	         + F_I*(gamma_I*e_FI*(A/(h_F +A)))  # births from infecteds, should = 0
	         - (m_FU * F_U)^2)*dt               # density dependent mortality
	# infected 1IH
  	# this should stay zero
	dF_I <- ((P*beta_I*F_U) 
	         - (d_FI*F_I)                       # deaths
	         - (m_FI * F_I)^2)*dt               # density dependent mortality
	# uninfected 2IH
	dS_U <- ((r_SU*S_U)                                              # births
	         - (D_U * e_DU * alpha_DUSU + D_I*e_DI*S_U*alpha_DISU)   # consumption by uninfected and infected predators, no natural mortality 
	         + r_SI * S_I                                            # births of infecteds, should = 0
	         - F_I * S_U * c)* dt                                    # loss due to movement to infected class
	# infected 2IH
	  # should stay zero
	dS_I <- ((F_I * S_U * c) -                                          # infection
	           (D_U * e_DU * alpha_DUSI + D_I * e_DI * alpha_DISI))*dt  # consumption by uninfected and infected predators
	
	A.simu[i] <- A + dA
	FU.simu[i] <- F_U + dF_U
	FI.simu[i] <- F_I + dF_I
	SU.simu[i] <- S_U + dS_U
	SI.simu[i] <- S_I + dS_I
}

# plot

plot(tset, A.simu,type='l',las=1,lwd=2,ylab='Population Size',xlab='Time', col=Acol,ylim=c(-0,100))
lines(tset, FU.simu, type = 'l', las = 1, lwd = 2, col = Fcol)
lines(tset, FI.simu, type = 'l', las = 1, lty = 3, col = Fcol)
lines(tset, SU.simu, type = 'l', las = 1, lwd = 2, col = Scol)
lines(tset, SI.simu, type = 'l', las = 1, lty = 3, col = Scol)
legend(x=70,y=90,legend=c('Algae', '1st Intermediate Host'),lwd=2,col=c(Acol, Fcol))
```

OK, insane exponential growth of 2IH/fish, looks crazy... but probably accurate since no natural mortality

### Let's add in definitive hosts, birds
This should start to make things interesting, at least for fish 2IHs
I should expect to see a typical type 1 functional response to prey from pred, so cycling and no equilibrium
$$
\begin{align}
\frac{d D_U}{dt} &= D_U(S_U e_{DU} \alpha_{DUSU} - d_{DU} - S_I e_{DU} \alpha_{DUSI}) + D_I(S_U e_{DISU} + S_I e_{DI} \alpha_{DISI})\\
\newline
\frac{d D_I}{dt} &= D_i S_I e_{DU} \alpha_{DUSI} - d_{DI}\\
\end{align}
$$

```{r algae_1IH_2IH_DH}
# we're not adding in parasites yet so
P <- 0
c <- 0
r_SI <- 0 # infected reproductive rate is zero


# set up empty vectors and initial values
A.simu <- NaN*tset; A.simu[1] <- 1
FI.simu <- NaN*tset; FI.simu[1] <- 0 # no parasites yet so this is 0
FU.simu <- NaN*tset; FU.simu[1] <- 1
SI.simu <- NaN*tset; SI.simu[1] <- 0 # no parasites yet so this is 0
SU.simu <- NaN*tset; SU.simu[1] <- 1
DI.simu <- NaN*tset; DI.simu[1] <- 0 # no parasites yet so this is 0
DU.simu <- NaN*tset; DU.simu[1] <- 1


# for loop
for(i in 2:length(tset)){
	dt <- tset[i]-tset[i-1]
	A <- A.simu[i-1]
	F_U <- FU.simu[i-1]
	F_I <- FI.simu[i-1]
	S_U <- SU.simu[i-1]
	S_I <- SI.simu[i-1]
	D_U <- DU.simu[i-1]
	D_I <- DI.simu[i-1]
	# algae
	dA <- (inp - A*(d_A + gamma_I*F_I + gamma_U*F_U))*dt
	# uninfected 1IH
	dF_U <- (F_U*(gamma_U*e_FU*(A/(h_F +A)) - d_FU)  # grazing * conversion efficiency * algae pop/handling time - death
	         - (P*beta_I*F_U)                   # parasites (0) * infection * uninfected pop, should = 0
	         + F_I*(gamma_I*e_FI*(A/(h_F +A)))  # births from infecteds, should = 0
	         - (m_FU * F_U)^2)*dt               # density dependent mortality  
	# infected 1IH
  	# this should stay zero
	dF_I <- ((P*beta_I*F_U) 
	         - (d_FI*F_I)                 # deaths
	         - (m_FI * F_I)^2)*dt         # density dependent mortality
	# uninfected 2IH
	dS_U <- ((r_SU*S_U)                                              # births
	         - S_U*(D_U * e_DU * alpha_DUSU + D_I*e_DI*alpha_DISU)   # consumption by uninfected and infected predators 
	         + (r_SI * S_I)                                          # births of infecteds, should = 0 because SI is 0
	         - (F_I * S_U * c))* dt                                  # loss due to movement to infected class, should = 0
	# infected 2IH
	  # should stay zero
	dS_I <- ((F_I * S_U * c) -                                          # infection
	           S_I*(D_U * e_DU * alpha_DUSI + D_I * e_DI * alpha_DISI))*dt  # consumption by uninfected and infected predators
	
	# uninfected DH
	dD_U <- (D_U *((S_U * e_DU * alpha_DUSU) - d_DU - (S_I * e_DU * alpha_DUSI)) # births - death - infection
	         + D_I*((S_U * e_DI * alpha_DISU) + (S_I * e_DI * alpha_DISI)))* dt  # births by infected class
	
	# infected DH
	  # should stay zero
	dD_I <- (D_I *((S_I * e_DU * alpha_DUSI) - d_DI))* dt  # new infections from eating infected 2IH - death (not sure how conversion rate fits in here -- maybe should be mass action w/o e)
	A.simu[i] <- A + dA
	FU.simu[i] <- F_U + dF_U
	FI.simu[i] <- F_I + dF_I
	SU.simu[i] <- S_U + dS_U
	SI.simu[i] <- S_I + dS_I
	DU.simu[i] <- D_U + dD_U
	DI.simu[i] <- D_I + dD_I
}

# plot

plot(tset, A.simu,type='l',las=1,lwd=2,ylab='Population Size',xlab='Time', col=Acol,ylim=c(0,10), xlim = c(0, 1000))
lines(tset, FU.simu, type = 'l', las = 1, lwd = 2, col = Fcol)
lines(tset, FI.simu, type = 'l', las = 1, lty = 3, col = Fcol)
lines(tset, SU.simu, type = 'l', las = 1, lwd = 2, col = Scol)
lines(tset, SI.simu, type = 'l', las = 1, lty = 3, col = Scol)
lines(tset, DU.simu, type = 'l', las = 1, lwd = 2, col = Dcol)
lines(tset, DI.simu, type = 'l', las = 1, lty = 3, col = Dcol)
legend(x=70,y=7,legend=c('Algae', '1st Intermediate Host','2nd Intermediate Host','Definitive Host'),lwd=2,col=c(Acol, Fcol,Scol,Dcol))
```

Looks ok, populations are really small, will deal with that later

## NOW LETS ADD PARASITES
### Adding Parasite Eggs
Using this Equation
$$
\begin{align}
\newline
\frac{d P}{dt} &=  r_P D_I - P(d_P + \beta_I F_I + \beta_U F_U) \\
\end{align}
$$
###Let's add in cercariae to infect the second intermediate hosts  
This is the equation. 
$$
\frac{dC}{dt} = F_I \gamma_I e_C (\frac{A}{h_F + A}) - d_CC - C(S_I + S_U)c
$$
The equlibrium value is then:
$$
C* = \frac {A F_I \gamma_I e_C}{(A + h_F)(d_C + c(S_I + S_U))}
$$
So we will set C* and add it into the previous equations.
```{r}
# Cstar <- (A * F_I * gamma_I * e_C)/((A + h_F)*(d_C + c *(S_I + S_U)))

# i have no idea how to plug this into the loop so i'm just going to include the equation below
```

Just kidding, we're not doing that. Inserting dC/dt to the system instead
```{r algae_1IH_2IH_DH_Psites}

# set up empty vectors and initial values
A.simu <- NaN*tset; A.simu[1] <- 1
FI.simu <- NaN*tset; FI.simu[1] <- 1 # lets see what happens if this starts at zero
FU.simu <- NaN*tset; FU.simu[1] <- 1
C.simu <- NaN*tset; C.simu[1] <- 0
SI.simu <- NaN*tset; SI.simu[1] <- 1 # lets see what happens if this starts at zero
SU.simu <- NaN*tset; SU.simu[1] <- 1
DI.simu <- NaN*tset; DI.simu[1] <- 1 # lets see what happens if this starts at zero
DU.simu <- NaN*tset; DU.simu[1] <- 1
P.simu <- NaN*tset; P.simu[1] <- 1

# for loop
for(i in 2:length(tset)){
	dt <- tset[i]-tset[i-1]
	A <- A.simu[i-1]    # maybe subsume later
	F_U <- FU.simu[i-1]
	F_I <- FI.simu[i-1]
	C <- C.simu[i-1]    # maybe subsume later
	S_U <- SU.simu[i-1]
	S_I <- SI.simu[i-1]
	D_U <- DU.simu[i-1]
	D_I <- DI.simu[i-1]
	P <- P.simu[i-1]

	# algae
	dA <- (inp - A*(d_A + gamma_I*F_I + gamma_U*F_U))*dt
	# uninfected 1IH
	dF_U <- (F_U*(gamma_U*e_FU*(A/(h_F +A)) - d_FU)  # grazing * conversion efficiency * algae pop/handling time - death
	         - (P*beta_I*F_U)                   # parasites (0) * infection * uninfected pop, should = 0
	         + F_I*(gamma_I*e_FI*(A/(h_F +A)))  # births from infecteds, should = 0
	         - (m_FU * F_U)^2)*dt               # density dependent mortality

	# infected 1IH
	dF_I <- ((P*beta_I*F_U) 
	         - (d_FI*F_I)                       # deaths
	         - (m_FI * F_I)^2)*dt               # density dependent mortality
	# cercariae
	# dC/dt = Fi g e A/(h+A) - m C - C (Si + Su)c
	dC <- (F_I * gamma_I * e_C * (A/(h_F + A)) # new cercariae made of algae via snails
	       - (d_C * C)                         # death of cercariae                       
	       - C* (S_I + S_U)*c)*dt              # uptake of cercariae into 2IH at a rate c
	# uninfected 2IH
	dS_U <- ((r_SU*S_U)                                              # births
	         - S_U*(D_U * e_DU * alpha_DUSU + D_I*e_DI*alpha_DISU)   # consumption by uninfected and infected predators 
	         + (r_SI * S_I)                                            # births of infecteds, should = 0 because SI is 0
	         - (F_I * S_U * c))* dt                                    # loss due to movement to infected class
	# infected 2IH
	dS_I <- ((C * S_U * c) -                                           # infection
	           S_I*(D_U * e_DU * alpha_DUSI + D_I * e_DI * alpha_DISI))*dt  # consumption by uninfected and infected predators
	
	# uninfected DH
	dD_U <- (D_U *((S_U * e_DU * alpha_DUSU) - d_DU - (S_I * e_DU * alpha_DUSI)) # births - death - infection
	         + D_I*((S_U * e_DI * alpha_DISU) + (S_I * e_DI * alpha_DISI)))* dt  # births by infected class
	
	# infected DH
	  # should stay zero
	dD_I <- ((D_U *(S_I * e_DU * alpha_DUSI)) - (D_I*d_DI))* dt  # new infections from eating infected 2IH - death (not sure how conversion rate fits in here -- maybe should be mass action w/o e)
	# parasite eggs
	dP <- ((r_P * D_I) - P * (d_P + beta_I * F_I + beta_U * F_U)) * dt
	
	A.simu[i] <- A + dA
	FU.simu[i] <- F_U + dF_U
	FI.simu[i] <- F_I + dF_I
	C.simu[i] <- C + dC
	SU.simu[i] <- S_U + dS_U
	SI.simu[i] <- S_I + dS_I
	DU.simu[i] <- D_U + dD_U
	DI.simu[i] <- D_I + dD_I
	P.simu[i] <- P + dP

}

# plot

plot(tset, A.simu,type='l',las=1,lwd=2,ylab='Population Size',xlab='Time', col=Acol,ylim=c(0,20), xlim = c(0, 1000))
lines(tset, FU.simu, type = 'l', las = 1, lwd = 2, col = Fcol)
lines(tset, FI.simu, type = 'l', las = 1, lty = 3, col = Fcol)
lines(tset, SU.simu, type = 'l', las = 1, lwd = 2, col = Scol)
lines(tset, SI.simu, type = 'l', las = 1, lty = 3, col = Scol)
lines(tset, DU.simu, type = 'l', las = 1, lwd = 2, col = Dcol)
lines(tset, DI.simu, type = 'l', las = 1, lty = 3, col = Dcol)
lines(tset, P.simu, type = 'l', las = 1, lwd = 2, col = Pcol)
lines(tset, C.simu, type = 'l', las = 1, lty = 2, col = "black")
legend("topright",legend=c('Algae', '1st Intermediate Host','2nd Intermediate Host','Definitive Host', 'Parasite Eggs', 'Cercariae'),lwd=2,col=c(Acol, Fcol,Scol,Dcol,Pcol,"black"))
```

Ugh. This is a mess. Where are my parasites going? Parameters!








