---
title: "Final Time Series"
author: "Zoe Zilz"
date: "5/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


```{r set parameters etc, include=FALSE}
#source("final_paramvals.R") 
source("WORKING_paramvals.R") # continually edited as chunks of the model are added# continually edited as chunks of the model are added

#source("build_statevars0.R")

tset <- seq(from=0, to=1000, length.out = 5000) #

# color for algae
Acol <- "green"
# color for 1IH snails
Fcol <- "orange"
# color for 2IH fish
Scol <- "purple"
# color for DH birbs
Dcol <- "blue"
# color for parasite eggs
Pcol <- "red"
```
Model:
$$
\begin{align}
\newline
Algae:\\
\frac{d A}{dt}  &= i - \left(\gamma_I F_I(\frac{A}{h_F+A})  + \gamma_U F_U(\frac{A}{h_F+A})\right)- Ad_A \\
\newline
First\ Intermediate\ Hosts\ (Snails):\\
\frac{d F_U}{dt} &= F_U (\gamma_U e_{FI}\frac{A}{h_F+A}) + F_I(\gamma_I e_{FI}\frac{A}{h_F+A}) -P\beta_I F_U - m{F_U}^2\\
\newline
\frac{d F_I}{dt} &= P\beta_I F_U - m{F_I}^2\\
\newline
Cercariae:\\
\frac{dC}{dt} &= F_I \gamma_I e_C (\frac{A}{h_F + A}) - d_CC - C(S_I + S_U)c \\
\newline
Second\ Intermediate\ Hosts\ (Fish):\\
\frac{d S_U}{dt} &= r_{SU}S_U + r_{SI} S_I - \left((D_U + D_I) * \gamma_D \frac{S_U}{h_D + (V + S_U + \alpha S_I)}\right)  - C S_U c - S_U m_{SU} \\
\newline
\frac{d S_I}{dt} &= C S_U c - \left((D_U + D_I) * \gamma_D \frac{\alpha S_I}{h_D + (V + S_U + \alpha S_I)}\right) - SI_I m_{SI}\\
\newline
Definitive\ Hosts\ (Birds):\\
\frac{d D_U}{dt} &= (D_U + D_I)\left(\gamma e_{DU} \frac{V +S_U + \alpha S_I}{h_D + (V + S_U + \alpha S_I)} \right) - D_U \frac{\gamma_D \alpha S_I}{h_D + (V + S_U + \alpha S_I)} - D_U d\\
\newline
\frac{d D_I}{dt} &= D_U \frac{\gamma_D \alpha S_I}{h + (V + S_U + \alpha S_I)} - D_I d\\
\newline
Parasite\ Eggs:
\newline
\frac{d P}{dt} &=  r_P D_I - P(d_P + \beta_I F_I + \beta_U F_U) \\
\end{align}
$$
## NOTES:
switch to mortality rate instead of logistic growth instead (for fish) *done*

need ratio of v/v+h needs to be less than mortality of DH
v < dh/(1-d)

do i need to incorporate the full handling time in denomenator with populations of all prey items (in denom) for the consumption part of the prey equation, while only having the prey population in the numerator??? *Yes*

```{r for loop}
# set up empty vectors and initial values
A.simu <- NaN*tset; A.simu[1] <- 1
FI.simu <- NaN*tset; FI.simu[1] <- 0 # parasites off
FU.simu <- NaN*tset; FU.simu[1] <- 1
C.simu <- NaN*tset; C.simu[1] <- 0
SI.simu <- NaN*tset; SI.simu[1] <- 0 # parasites off
SU.simu <- NaN*tset; SU.simu[1] <- 1
DI.simu <- NaN*tset; DI.simu[1] <- 0 # parasites off
DU.simu <- NaN*tset; DU.simu[1] <- 1
P.simu <- NaN*tset; P.simu[1] <- 1 # toggle on and off to see system without parasites

# for loop
for(i in 2:length(tset)){
  dt <- tset[i]-tset[i-1]
  A <- A.simu[i-1]    # maybe subsume later
  F_U <- FU.simu[i-1]
  F_I <- FI.simu[i-1]
  C <- C.simu[i-1]    # maybe subsume later
  S_U <- SU.simu[i-1]
  S_I <- SI.simu[i-1]
  D_U <- DU.simu[i-1]
  D_I <- DI.simu[i-1]
  P <- P.simu[i-1]
  
  # algae
  dA <- (inp - A*d_A                    # input - death
         - gamma_I*F_I*(A/(h_F +A))     # - grazing rate x handling time infected
         - gamma_U*F_U*(A/(h_F +A)))*dt # - grazing rate x handling time uninfected
  # uninfected 1IH
  dF_U <- (F_U*(gamma_U*e_FU*(A/(h_F +A)))  # grazing * conversion efficiency * algae pop/handling time (decreases with prey density) - REMOVED: death (-d_FU)
           - (P*beta_I*F_U)                   # parasites * infection * uninfected pop
           + F_I*(gamma_I*e_FI*(A/(h_F + A))) # births from infecteds
           - (m_FU * F_U)^2)*dt               # density dependent mortality
  
  # infected 1IH
  dF_I <- ((P*beta_I*F_U) 
           #- (d_FI*F_I)                      # REMOVED BACKGROUND MORTALITY, too much death
           - (m_FI * F_I)^2)*dt               # density dependent mortality
  # cercariae
  # dC/dt = Fi g e A/(h+A) - m C - C (Si + Su)c
  dC <- (F_I * gamma_I * e_C * (A/(h_F + A))                         # new cercariae made of algae via snails
         - (d_C * C)                                                 # death of cercariae                       
         - C* (S_I + S_U)*c)*dt                                      # uptake of cercariae into 2IH at a rate c
  
  
  # uninfected 2IH
  dS_U <- (
    (r_SU*S_U)                              
    -((D_U) * (gamma_D * ((S_U)/(h_D + (v + S_U + alpha* S_I)))))       # added in all prey items to denom
    -(D_I * gamma_D * ((S_U)/(h_D + (v + S_U + alpha* S_I))))            # un-simplified equation, multiplied it out
    + (r_SI * S_I)                   # births from infecteds, off right now
    - (F_I * S_U * c)                # loss due to movement to infected class, also off right now
    - S_U * m_SU
  )* dt                              
  
  
  
  # infected 2IH ## should be zero right now
  dS_I <- ((C * S_U * c)                                                               # infection
           - ((D_U + D_I) * (gamma_D * ((alpha* S_I)/(h_D + (v + S_U + alpha* S_I))))) # type II consumption by uninfected+infected predators 
           - S_I * m_SI
           )*dt  
  
  
  
  
  # uninfected DH
  dD_U <- ((D_U + D_I) * (gamma_D * e_DU * ((v + S_U + alpha* S_I)/(h_D + (v + S_U + alpha * S_I)))) # births due to consumption of infected and uninfected prey, type II functional response
           - (D_U * d_DU)                                                                 # minus deaths
           - D_U*((gamma_D*alpha*S_I)/(h_D + (v+S_U + alpha * S_I))) # minus movement to infected class
  )*dt                               
  
  
  
  # infected DH ## should be zero right now
  dD_I <- ((D_U *((gamma_D * alpha * S_I)/(h_D + (v + S_U + alpha * S_I)))) # new infections from eating infected 2IH
           - (D_I*d_DI)                                                     # minus death
  )* dt                                         
  
  # parasite eggs
  dP <- ((r_P * D_I) - P * (d_P + beta_I * F_I + beta_U * F_U)) * dt
  
  A.simu[i] <- A + dA
  FU.simu[i] <- F_U + dF_U
  FI.simu[i] <- F_I + dF_I
  C.simu[i] <- C + dC
  SU.simu[i] <- S_U + dS_U
  SI.simu[i] <- S_I + dS_I
  DU.simu[i] <- D_U + dD_U
  DI.simu[i] <- D_I + dD_I
  P.simu[i] <- P + dP
  
}
```

Plot Time Series
```{r plot time series}
# plot
# parasites are off so parasite/infected time series are commented out
# legend gets in the way so its commented out

#png("withparasites.png")
plot(tset, A.simu,type='l',las=1,lwd=2,ylab='Population Size',xlab='Time', col=Acol,ylim=c(0,100), xlim = c(0, 1000))
lines(tset, FU.simu, type = 'l', las = 1, lwd = 2, col = Fcol)
lines(tset, FI.simu, type = 'l', las = 1, lty = 3, col = Fcol)
lines(tset, SU.simu, type = 'l', las = 1, lwd = 2, col = Scol)
lines(tset, SI.simu, type = 'l', las = 1, lty = 3, col = Scol)
lines(tset, DU.simu, type = 'l', las = 1, lwd = 2, col = Dcol)
abline(h = 0)
lines(tset, DI.simu, type = 'l', las = 1, lty = 3, col = Dcol)
lines(tset, P.simu, type = 'l', las = 1, lwd = 2, col = Pcol)
lines(tset, C.simu, type = 'l', las = 1, lty = 2, col = "black")
#legend("topright",legend=c('Algae', '1st Intermediate Host','2nd Intermediate Host','Definitive Host', 'Parasite Eggs', 'Cercariae'),lwd=2,col=c(Acol, Fcol,Scol,Dcol,Pcol,"black"))
#dev.off()

```

Repeat above with initial conditions already at equilibrium, just to see what happens
```{r time series at eq}

A.simu2 <- NaN*tset; A.simu2[1] <- A.simu[length(tset)]
FI.simu2 <- NaN*tset; FI.simu2[1] <- FI.simu[length(tset)] # parasites off
FU.simu2 <- NaN*tset; FU.simu2[1] <- FU.simu[length(tset)]
C.simu2 <- NaN*tset; C.simu2[1] <- C.simu[length(tset)]
SI.simu2 <- NaN*tset; SI.simu2[1] <- SI.simu[length(tset)] # parasites off
SU.simu2 <- NaN*tset; SU.simu2[1] <- SU.simu[length(tset)]
DI.simu2 <- NaN*tset; DI.simu2[1] <- DI.simu[length(tset)] # parasites off
DU.simu2 <- NaN*tset; DU.simu2[1] <- DU.simu[length(tset)]
P.simu2 <- NaN*tset; P.simu2[1] <- P.simu[length(tset)] # toggle on and off to see system without parasites

# for loop
for(i in 2:length(tset)){
  dt <- tset[i]-tset[i-1]
  A <- A.simu2[i-1]    # maybe subsume later
  F_U <- FU.simu2[i-1]
  F_I <- FI.simu2[i-1]
  C <- C.simu2[i-1]    # maybe subsume later
  S_U <- SU.simu2[i-1]
  S_I <- SI.simu2[i-1]
  D_U <- DU.simu2[i-1]
  D_I <- DI.simu2[i-1]
  P <- P.simu2[i-1]
  
  # algae
  dA <- (inp - A*d_A                    # input - death
         - gamma_I*F_I*(A/(h_F +A))     # - grazing rate x handling time infected
         - gamma_U*F_U*(A/(h_F +A)))*dt # - grazing rate x handling time uninfected
  # uninfected 1IH
  dF_U <- (F_U*(gamma_U*e_FU*(A/(h_F +A)))  # grazing * conversion efficiency * algae pop/handling time (decreases with prey density) - REMOVED: death (-d_FU)
           - (P*beta_I*F_U)                   # parasites * infection * uninfected pop
           + F_I*(gamma_I*e_FI*(A/(h_F + A))) # births from infecteds
           - (m_FU * F_U)^2)*dt               # density dependent mortality
  
  # infected 1IH
  dF_I <- ((P*beta_I*F_U) 
           #- (d_FI*F_I)                      # REMOVED BACKGROUND MORTALITY, too much death
           - (m_FI * F_I)^2)*dt               # density dependent mortality
  # cercariae
  # dC/dt = Fi g e A/(h+A) - m C - C (Si + Su)c
  dC <- (F_I * gamma_I * e_C * (A/(h_F + A))                         # new cercariae made of algae via snails
         - (d_C * C)                                                 # death of cercariae                       
         - C* (S_I + S_U)*c)*dt                                      # uptake of cercariae into 2IH at a rate c
  
  
  # uninfected 2IH
  dS_U <- (
    (r_SU*S_U)                              
    -((D_U) * (gamma_D * ((S_U)/(h_D + (v + S_U + alpha* S_I)))))       # added in all prey items to denom
    -(D_I * gamma_D * ((S_U)/(h_D + (v + S_U + alpha* S_I))))            # un-simplified equation, multiplied it out
    + (r_SI * S_I)                   # births from infecteds, off right now
    - (F_I * S_U * c)                # loss due to movement to infected class, also off right now
    - S_U * m_SU
  )* dt                              
  
  
  
  # infected 2IH ## should be zero right now
  dS_I <- ((C * S_U * c)                                                               # infection
           - ((D_U + D_I) * (gamma_D * ((alpha* S_I)/(h_D + (v + S_U + alpha* S_I))))) # type II consumption by uninfected+infected predators 
           - S_I * m_SI
           )*dt  
  
  
  
  
  # uninfected DH
  dD_U <- ((D_U + D_I) * (gamma_D * e_DU * ((v + S_U + alpha* S_I)/(h_D + (v + S_U + alpha * S_I)))) # births due to consumption of infected and uninfected prey, type II functional response
           - (D_U * d_DU)                                                                 # minus deaths
           - D_U*((gamma_D*alpha*S_I)/(h_D + (v+S_U + alpha * S_I))) # minus movement to infected class
  )*dt                               
  
  
  
  # infected DH ## should be zero right now
  dD_I <- ((D_U *((gamma_D * alpha * S_I)/(h_D + (v + S_U + alpha * S_I)))) # new infections from eating infected 2IH
           - (D_I*d_DI)                                                     # minus death
  )* dt                                         
  
  # parasite eggs
  dP <- ((r_P * D_I) - P * (d_P + beta_I * F_I + beta_U * F_U)) * dt
  
  A.simu2[i] <- A + dA
  FU.simu2[i] <- F_U + dF_U
  FI.simu2[i] <- F_I + dF_I
  C.simu2[i] <- C + dC
  SU.simu2[i] <- S_U + dS_U
  SI.simu2[i] <- S_I + dS_I
  DU.simu2[i] <- D_U + dD_U
  DI.simu2[i] <- D_I + dD_I
  P.simu2[i] <- P + dP
  
}

# plot
# parasites are off so parasite/infected time series are commented out
# legend gets in the way so its commented out

#png("withparasites.png")
plot(tset, A.simu2,type='l',las=1,lwd=2,ylab='Population Size',xlab='Time', col=Acol,ylim=c(0,100), xlim = c(0, 1000))
lines(tset, FU.simu2, type = 'l', las = 1, lwd = 2, col = Fcol)
lines(tset, FI.simu2, type = 'l', las = 1, lty = 3, col = Fcol)
lines(tset, SU.simu2, type = 'l', las = 1, lwd = 2, col = Scol)
lines(tset, SI.simu2, type = 'l', las = 1, lty = 3, col = Scol)
lines(tset, DU.simu2, type = 'l', las = 1, lwd = 2, col = Dcol)
abline(h = 0)
lines(tset, DI.simu2, type = 'l', las = 1, lty = 3, col = Dcol)
lines(tset, P.simu2, type = 'l', las = 1, lwd = 2, col = Pcol)
lines(tset, C.simu2, type = 'l', las = 1, lty = 2, col = "black")
#legend("topright",legend=c('Algae', '1st Intermediate Host','2nd Intermediate Host','Definitive Host', 'Parasite Eggs', 'Cercariae'),lwd=2,col=c(Acol, Fcol,Scol,Dcol,Pcol,"black"))
#dev.off()
```
Everything flatlines, so I think this is good


###Now I'm going to play with parameter values and inital conditions
Holly says to increase overall abundance I can mess with death rates and conversion rates

```{r}
source("messingaround_paramvals.R")
tset <- seq(from=0, to=1000, length.out = 50000)


# set up empty vectors and initial values
A.simu <- NaN*tset; A.simu[1] <- 1
FI.simu <- NaN*tset; FI.simu[1] <- 1 #0 parasites off
FU.simu <- NaN*tset; FU.simu[1] <- 1
C.simu <- NaN*tset; C.simu[1] <- 0
SI.simu <- NaN*tset; SI.simu[1] <- 0 #0 parasites off
SU.simu <- NaN*tset; SU.simu[1] <- 1
DI.simu <- NaN*tset; DI.simu[1] <- 0 #0 parasites off
DU.simu <- NaN*tset; DU.simu[1] <- 1
P.simu <- NaN*tset; P.simu[1] <- 0 # toggle on and off to see system without parasites

# for loop
for(i in 2:length(tset)){
  dt <- tset[i]-tset[i-1]
  A <- A.simu[i-1]    # maybe subsume later
  F_U <- FU.simu[i-1]
  F_I <- FI.simu[i-1]
  C <- C.simu[i-1]    # maybe subsume later
  S_U <- SU.simu[i-1]
  S_I <- SI.simu[i-1]
  D_U <- DU.simu[i-1]
  D_I <- DI.simu[i-1]
  P <- P.simu[i-1]
  
  # algae
  dA <- (inp - A*d_A                    # input - death
         - gamma_I*F_I*(A/(h_F +A))     # - grazing rate x handling time infected
         - gamma_U*F_U*(A/(h_F +A)))*dt # - grazing rate x handling time uninfected
  # uninfected 1IH
  dF_U <- (F_U*(gamma_U*e_FU*(A/(h_F +A)))  # grazing * conversion efficiency * algae pop/handling time (decreases with prey density) - REMOVED: death (-d_FU)
           - (P*beta_I*F_U)                   # parasites * infection * uninfected pop
           + F_I*(gamma_I*e_FI*(A/(h_F + A))) # births from infecteds
           - (m_FU * F_U)^2)*dt               # density dependent mortality
  
  # infected 1IH
  dF_I <- ((P*beta_I*F_U) 
           #- (d_FI*F_I)                      # REMOVED BACKGROUND MORTALITY, too much death
           - (m_FI * F_I)^2)*dt               # density dependent mortality
  # cercariae
  # dC/dt = Fi g e A/(h+A) - m C - C (Si + Su)c
  dC <- (F_I * gamma_I * e_C * (A/(h_F + A))                         # new cercariae made of algae via snails
         - (d_C * C)                                                 # death of cercariae                       
         - C* (S_I + S_U)*c)*dt                                      # uptake of cercariae into 2IH at a rate c
  
  
  # uninfected 2IH
  dS_U <- (
    (r_SU*S_U)                              
    -((D_U) * (gamma_D * ((S_U)/(h_D + (v + S_U + alpha* S_I)))))       # added in all prey items to denom
    -(D_I * gamma_D * ((S_U)/(h_D + (v + S_U + alpha* S_I))))            # un-simplified equation, multiplied it out
    + (r_SI * S_I)                   # births from infecteds, off right now
    - (F_I * S_U * c)                # loss due to movement to infected class, also off right now
    - S_U * m_SU
  )* dt                              
  
  
  
  # infected 2IH ## should be zero right now
  dS_I <- ((C * S_U * c)                                                               # infection
           - ((D_U + D_I) * (gamma_D * ((alpha* S_I)/(h_D + (v + S_U + alpha* S_I))))) # type II consumption by uninfected+infected predators 
           - S_I * m_SI
           )*dt  
  
  
  
  
  # uninfected DH
  dD_U <- ((D_U + D_I) * (gamma_D * e_DU * ((v + S_U + alpha* S_I)/(h_D + (v + S_U + alpha * S_I)))) # births due to consumption of infected and uninfected prey, type II functional response
           - (D_U * d_DU)                                                                 # minus deaths
           - D_U*((gamma_D*alpha*S_I)/(h_D + (v+S_U + alpha * S_I))) # minus movement to infected class
  )*dt                               
  
  
  
  # infected DH ## should be zero right now
  dD_I <- ((D_U *((gamma_D * alpha * S_I)/(h_D + (v + S_U + alpha * S_I)))) # new infections from eating infected 2IH
           - (D_I*d_DI)                                                     # minus death
  )* dt                                         
  
  # parasite eggs
  dP <- ((r_P * D_I) - P * (d_P + beta_I * F_I + beta_U * F_U)) * dt
  
  A.simu[i] <- A + dA
  FU.simu[i] <- F_U + dF_U
  FI.simu[i] <- F_I + dF_I
  C.simu[i] <- C + dC
  SU.simu[i] <- S_U + dS_U
  SI.simu[i] <- S_I + dS_I
  DU.simu[i] <- D_U + dD_U
  DI.simu[i] <- D_I + dD_I
  P.simu[i] <- P + dP
}

# plot
# legend gets in the way so its commented out
```

```{r}
#png("ploot.png")
plot(tset, A.simu,type='l',las=1,lwd=2,ylab='Population Size',xlab='Time', col=Acol,ylim=c(0,50), xlim = c(0, 1000))
lines(tset, FU.simu, type = 'l', las = 1, lwd = 2, col = Fcol)
lines(tset, FI.simu, type = 'l', las = 1, lty = 3, col = Fcol)
lines(tset, SU.simu, type = 'l', las = 1, lwd = 2, col = Scol)
lines(tset, SI.simu, type = 'l', las = 1, lty = 3, col = Scol)
lines(tset, DU.simu, type = 'l', las = 1, lwd = 2, col = Dcol)
abline(h = 0)
lines(tset, DI.simu, type = 'l', las = 1, lty = 3, col = Dcol)
lines(tset, P.simu, type = 'l', las = 1, lwd = 2, col = Pcol)
lines(tset, C.simu, type = 'l', las = 1, lty = 2, col = "black")
#legend("topright",legend=c('Algae', '1st Intermediate Host','2nd Intermediate Host','Definitive Host', 'Parasite Eggs', 'Cercariae'),lwd=2,col=c(Acol, Fcol,Scol,Dcol,Pcol,"black"))
#dev.off()
```


